Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-08-02 manuscripts/reply/reply-main.tex
More editing of reply ms.

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method is very sensitive to prior assumptions and often
supports shared divergences across taxa that diverged randomly over broad time
periods, to which \citet{Hickerson2013} responded.
\citet{Oaks2012} and \citet{Hickerson2013} agree on the fundamental
methodological point about the model selection performed in \msb:
\begin{itemize}
   \item The use of vague priors on the divergence-time parameters can lead to
       spurious support for models with few divergence events shared across
       taxa. Thus, the primary inference enabled by the approach is very
       sensitive to the priors on divergence times.
\end{itemize}
However, the two papers suggest alternative mechanisms by which the priors on
divergence times cause this behavior:
\begin{enumerate}[label=\textsl{\textbf{Hypothesis \arabic*)}},ref=\arabic*,align=left]
    \item \textsl{\textbf{Numerical approximation error}} \citep{Hickerson2013}:
        Under broad uniform priors, the rejection algorithm implemented in \msb
        is unable to adequately sample the space of the models within
        reasonable computational time, which leads to bias toward models with
        fewer divergence-time parameters because they are better sampled.
        \label{hypError}
    % \item \citet{Hickerson2013} suggest that under broad uniform priors, the
    %     rejection algorithm implemented in \msb is unable to adequately sample
    %     the space of the models within reasonable computational time, which
    %     leads to bias toward models with fewer divergence-time parameters
    %     because they are better sampled.
    \item \textsl{\textbf{Strongly weighted marginal likelihoods}}
        \citep{Oaks2012}:
        The uniform priors on divergence times lead to very small marginal
        likelihoods (and thus posterior probabilities) of models with many
        divergence-time parameters.  The likelihood of these models is
        ``averaged'' over a much greater parameter space in which there is a
        large amount of prior weight and small probability of producing the
        data.  \citep{Jeffreys1939,Lindley1957}. \label{hypML}
    % \item \citet{Oaks2012} suggest the cause is more fundamental; the uniform
    %     priors on divergence times simply lead to very small marginal
    %     likelihoods of models with many divergence-time parameters. These
    %     models are ``averaged'' over much greater parameter space with large
    %     prior weight and small likelihood \citep{Jeffreys1939,Lindley1957}.
\end{enumerate}
In Hypothesis~\ref{hypError}, the problem is numerical approximation error due to
insufficient computation;
given data from randomly diverged taxa, the exact posterior supports a model
with many divergence-time parameters, but we are unable to accurately
approximate this posterior.
In Hypothesis~\ref{hypML}, the problem is more fundamental; given data from
randomly diverged taxa, the exact posterior supports a model with a small
number of simultaneous divergences.
% In the latter case, numerical error is not the issue, because the exact
% posterior supports simultaneous divergences given data from randomly diverged
% taxa.
I.e., given the assumption that divergence times are broadly and uniformly
distributed, the exact posterior from Bayes' rule leads us to the wrong
conclusion about evolutionary history.
% I.e., this is the correct (albeit biogeographically misleading) answer given
% the prior assumption that divergence times are broadly and uniformly
% distributed.
Such posterior support for simultaneous divergence, even if ``correct'' from
the perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.

@@

%%
\section{The causes of support for models of co-divergence}
To determine how best to improve the behavior of \msb, it is important to
determine the mechanism by which broad uniform priors cause support for
clustered models of divergence.
It is well established that vague priors can be problematic in Bayesian model
selection.
Models that integrate over more parameter space characterized by low
probability of producing the data and relatively high prior density will have
smaller marginal likelihoods \citep{Jeffreys1939,Lindley1957}.
Given the uniformly distributed priors on divergence times employed in \msb,
the likelihood of models with more divergence parameters will be ``averaged''
over much greater parameter space, all with equal prior weight, and much of it
with small likelihood.
In light of this fundamental statistical issue, it is not surprising that the
method tends to support simple models.

However, \citet{Hickerson2013} conclude that the bias is caused by
numerical approximation error due to insufficient computation.
They argue the widest of the three priors used by \citet{Oaks2012} would
infrequently produce random samples with many independent population divergence
times as recent as the estimated gene divergence times presented in
\citet{Oaks2012}.
However, this argument assumes the gene divergence times presented in
\citet{Oaks2012} were estimated without error.
These estimates were intended to provide only a rough comparison of the
gene divergence times across the 22 taxa.
These analyses assumed an arbitrary strict per-site rate of $2\e{-8}$ mutations
per generation for all taxa, and are, of course, subject to estimation error.
Furthermore, the branch-length units of the gene trees are in millions of
years, whereas the divergence-time prior of \msb is in generations, thus
\citet{Hickerson2013} make the implicit assumption that all 22 Philippine taxa
have a generation time of one year.
The argument of \citet{Hickerson2013} that divergence-time estimates of
\citet{Oaks2012} should set an upper bound on the prior for divergence times
seems questionable, especially given our findings presented above regarding the
behavior of \msb when empirically informed uniform priors are employed.

Even if we assume the arbitrary strict clock is correct, gene divergence times
were estimated without error, and all 22 taxa have one-year generation times,
Hickerson et al.'s
(\citeyear{Hickerson2013}) argument only applies to one of the \emph{three}
priors used by \citet{Oaks2012}.
Under these assumptions, the narrowest prior on divergence times used by
\citet{Oaks2012} ($U(0, 5)$) closely mirrors the range of estimates of
gene-divergence times (0--5 million years ago).
Applying Hickerson et al.'s (\citeyear{Hickerson2013}) sampling-probability
argument demonstrates this prior is densely populated with samples with large
numbers of divergence parameters with values younger than the estimated gene
divergence estimates.
Thus, if insufficient prior sampling is to blame for the bias, it should be
much reduced under the narrow prior on \divt{}.
However, the magnitude of the bias is very similar across all three priors
explored by \citet{Oaks2012}.
\citet{Hickerson2013} point out a case where the narrowest prior performs
slightly better (panel L of Figures S32, S37, and S38 of \citet{Oaks2012}).
However, it is important to note that these results suffered from a bug
in \msb, and there are many cases after \citet{Oaks2012} corrected the 
bug where the narrowest prior performs slightly worse (see panels D--J of
Figures 3 and S12).

To disentangle whether model likelihoods or numerical approximation error is to
blame for the method's erroneous support for simple models, we must look at the
different predictions made by these two phenomena.
For example, numerical error due to insufficient prior sampling should create
large variance among posterior estimates and cause analyses to be highly
sensitive to the number of samples drawn from the prior.
Furthermore, if insufficient prior sampling is \emph{biasing} estimates toward
models with less parameter space, as suggested by \citet{Hickerson2013}, we
expect to see support for these models decrease as sampling from the prior
increases.
\citet{Oaks2012} did not see such sensitivity when they compared prior sample
sizes of $2\e6$, $5\e6$, and $10^7$.

To explore this prediction further, we repeat the analysis of the Philippines
dataset under the intermediate prior used by \citet{Oaks2012} ($\divt{} \sim
U(0, 10)$, $\meanDescendantTheta{} \sim (0.0005, 0.04)$, $\ancestralTheta{}
\sim (0.0005, 0.02)$), using a very large prior sample size of $10^8$.
When we look at the trace of the estimates of the dispersion index of
divergence times (\vmratio{}) as the prior samples accumulate
(Figure~S\ref{figSamplingError}) we do not see the trend predicted by
\citet{Hickerson2013}.
While approximation error is always present for any numerical analysis, it does
not appear to be playing a large role in the biases revealed by the results of
\citet{Oaks2012} or presented above.

A straightforward prediction if marginal likelihoods are causing the preference
for simple models is that the bias should disappear as the model generating the
data converges to the prior.
\citet{Oaks2012} tested this prediction by performing 100,000 simulations to
assess the model-choice behavior of \msb when the prior model is correct.
The results confirm the prediction:
\msb estimates the probability of the one-divergence model quite well when the
prior is correct (see Figure 4 of \citet{Oaks2012}).
We confirmed this same behavior for the model-averaging approach used by
\citet{Hickerson2013} (see SI text and Figure~S\ref{figValidationMCBehavior}).
These results are not clearly predicted if insufficient computation was causing
numerical error.
Even when the prior is correct, due to the discrete uniform prior on the number
of divergence events (\numt{}) implemented in \msb, models with larger numbers
of divergence-time parameters (and thus greater parameter space) will still be
far less densely sampled than those with fewer divergence events
\citep{Oaks2012}.
Thus, the results of the simulations of \citet{Oaks2012} are more consistent
with the fundamental sensitivity of marginal likelihoods to priors.

@@

\section{Improving inference of shared divergences}
In theory, the model-averaging approach of \citet{Hickerson2013} is appealing.
It leverages a great strength of Bayesian statistical procedures, namely the
ability to obtain marginalized estimates that incorporate uncertainty in
nuisance parameters.
However, when sampling over models with narrow-empirical and diffuse uniform
priors for a parameter that is expected to have a very non-uniform likelihood
density, models that exclude the true values of the parameters we aim to
estimate will often have the largest marginal likelihood.

The recommendations of \citet{Oaks2012} for mitigating the lack of robustness
of \msb are similar to those of \citet{Hickerson2013}, but avoid the
need for imposing an additional dimension of model choice.
\citet{Oaks2012} suggest that uniform priors may not be ideal for many
parameters of the \msb model, and recommend the use of probability
distributions from the exponential family.
If we look at the prior distribution on divergence times imposed by
the model-averaging approach of \citet{Hickerson2013} we see it is a mixture of
overlapping uniforms with lower limits of zero (Figure~S\ref{figMCTauPrior}).
This looks very much like an exponential distribution, except that in any state
of the model, all the divergence times are restricted to the hard
bounds of one of the uniform distributions.
Thus, it seems more appropriate to simply place a gamma prior (the exponential
being a special case) on divergence times.
This would capture the prior uncertainty that \citet{Hickerson2013} are
suggesting for divergence times (Figure~S\ref{figMCTauPrior}) while avoiding
costly model-averaging and the constraint that all divergence times must fall
within the hard bounds of the current model state.
It also would allow an investigator to place the majority of the prior density
in regions of parameter space they believe, \emph{a priori}, are most
plausible, but still capture uncertainty in the tails of distributions with low
density.
Indeed, \citet{Oaks2014dpp} has shown that the use of gamma distributions in
place of uniform priors improves the power of the method to detect temporal
variation in divergences and reduces erroneous support for clustered
divergences.

\section{Conclusions}
We demonstrate how the approximate Bayesian model-choice method implemented in
\msb can spuriously support models with less parameter space.
This is caused by the use of uniform priors on divergence times.
Uniform distributions necessitate the use of priors that place high density in
unlikely regions of parameter space, less the risk of excluding the true
divergence times \emph{a priori}.
These broad uniform priors reduce the marginal likelihoods of models with more
divergence-time parameters.
We show that the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} does not mitigate this bias, but rather causes it to
manifest by sampling predominantly from models that may exclude the true values
of the parameters.
Our results show that it is difficult to choose an uniformly distributed
prior on divergence times that is broad enough to confidently contain the true
values of parameters while being narrow enough to avoid erroneous support for
models with less parameter space.
More generally, it is important to carefully choose prior assumptions about
parameters in Bayesian model selection; these prior assumptions about
parameters strongly influence the posterior probabilities we seek to estimate.

The common inference of temporally clustered divergences \citep{Barber2010,
    Bell2012, Carnaval2009, Chan2011, Daza2010, Hickerson2006, Huang2011,
    Lawson2010, Leache2007, Plouviez2009, Stone2012, Voje2009}, when not
accompanied with the necessary analyses to assess the robustness and temporal
resolution of such results, should be treated with caution, because \msb has
been shown to erroneously infer clustered divergences over a range of prior
conditions.
Fortunately, \citet{Oaks2014dpp} has shown that alternative probability
distributions allow prior uncertainty to be accommodated while avoiding
excessive prior density in regions of low likelihood, which greatly improves
the ability to infer shared divergence histories.

>>>>

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method is very sensitive to prior assumptions and often
supports shared divergences across taxa that diverged randomly over broad time
periods, to which \citet{Hickerson2013} responded.
\citet{Oaks2012} and \citet{Hickerson2013} agree on the fundamental
methodological point about the model selection performed in \msb:
\begin{itemize}
   \item The use of vague priors on the divergence-time parameters can lead to
       spurious support for models with few divergence events shared across
       taxa. Thus, the primary inference enabled by the approach is very
       sensitive to the priors on divergence times.
\end{itemize}
However, the two papers suggest alternative mechanisms by which the priors on
divergence times cause this behavior:
\begin{enumerate}[label=\textsl{\textbf{Hypothesis \arabic*)}},ref=\arabic*,align=left]
    \item \textsl{\textbf{Numerical approximation error}} \citep{Hickerson2013}:
        Under broad uniform priors, the rejection algorithm implemented in \msb
        is unable to adequately sample the space of the models within
        reasonable computational time, which leads to bias toward models with
        fewer divergence-time parameters because they are better sampled.
        \label{hypError}
    \item \textsl{\textbf{Strongly weighted marginal likelihoods}}
        \citep{Oaks2012}:
        The uniform priors on divergence times lead to very small marginal
        likelihoods (and thus posterior probabilities) of models with many
        divergence-time parameters.  The likelihood of these models is
        ``averaged'' over a much greater parameter space in which there is a
        large amount of prior weight and small probability of producing the
        data.  \citep{Jeffreys1939,Lindley1957}. \label{hypML}
\end{enumerate}
In Hypothesis~\ref{hypError}, the problem is numerical approximation error due to
insufficient computation;
given data from randomly diverged taxa, the exact posterior supports a model
with many divergence-time parameters, but we are unable to accurately
approximate this posterior.
In Hypothesis~\ref{hypML}, the problem is more fundamental; given data from
randomly diverged taxa, the exact posterior supports a model with a small
number of simultaneous divergences.
I.e., given the assumption that divergence times are broadly and uniformly
distributed, the exact posterior from Bayes' rule leads us to the wrong
conclusion about evolutionary history.
Such posterior support for simultaneous divergence, even if ``correct'' from
the perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.

@@

%%
\section{The causes of support for models of co-divergence}
To determine how best to improve the behavior of \msb, it is important to
determine the mechanism by which broad uniform priors cause support for
clustered models of divergence.
It is well established that vague priors can be problematic in Bayesian model
selection.
Models that integrate over more parameter space characterized by low
probability of producing the data and relatively high prior density will have
smaller marginal likelihoods \citep{Jeffreys1939,Lindley1957}.
Given the uniformly distributed priors on divergence times employed in \msb,
the likelihood of models with more divergence parameters will be ``averaged''
over much greater parameter space, all with equal prior weight, and much of it
with small likelihood (Hypothesis~\ref{hypML}).
In light of this fundamental statistical issue, it is not surprising that the
method tends to support simple models.

However, \citet{Hickerson2013} conclude that the bias is caused by numerical
approximation error due to insufficient computation
(Hypothesis~\ref{hypError}).
They argue the widest of the three priors used by \citet{Oaks2012} would
infrequently produce random samples with many independent population divergence
times as recent as the estimated gene divergence times presented in
\citet{Oaks2012}.
However, this argument assumes the gene divergence times presented in
\citet{Oaks2012} were estimated without error.
These estimates were intended to provide only a rough comparison of the
gene divergence times across the 22 taxa.
These analyses assumed an arbitrary strict per-site rate of $2\e{-8}$ mutations
per generation for all taxa, and are, of course, subject to estimation error.
Furthermore, the branch-length units of the gene trees are in millions of
years, whereas the divergence-time prior of \msb is in generations, thus
\citet{Hickerson2013} make the implicit assumption that all 22 Philippine taxa
have a generation time of one year.
The argument of \citet{Hickerson2013} that these estimated gene divergence
times should be used to set the upper bound of the uniform prior on divergence
times for the \msb analysis of the same sequence data seems questionable,
especially given our findings presented above regarding the behavior of \msb
when empirically informed uniform priors are employed.

Even if we assume (1) the arbitrary strict clock is correct, (2) gene
divergence times were estimated without error, and (3) all 22 taxa have
one-year generation times, Hickerson et al.'s
(\citeyear{Hickerson2013}) argument only applies to one of the \emph{three}
priors used by \citet{Oaks2012}.
Under these assumptions, the narrowest prior on divergence times used by
\citet{Oaks2012} ($U(0, 5)$) closely mirrors the range of estimates of
gene divergence times (0--5 million years ago).
Applying Hickerson et al.'s (\citeyear{Hickerson2013}) sampling-probability
argument demonstrates this prior is densely populated with samples with large
numbers of divergence parameters with values younger than the estimated gene
divergence estimates.
Thus, if insufficient prior sampling (Hypothesis~\ref{hypError}) is to blame
for the bias, it should be much reduced under the narrow prior on \divt{}.
However, the magnitude of the bias is very similar across all three priors
explored by \citet{Oaks2012}.
\citet{Hickerson2013} point out a case where the narrowest prior performs
slightly better (panel L of Figures S32, S37, and S38 of \citet{Oaks2012}).
However, it is important to note that these results suffered from a bug in
\msb, and after \citet{Oaks2012} corrected the bug, there are many cases where
the narrowest prior performs slightly worse (see panels D--J of Figures 3 and
S12).

To disentangle whether Hypothesis 1 or 2 is the primary cause the method's
erroneous support for simple models, we must look at the different predictions
made by these two phenomena.
For example, numerical error due to insufficient prior sampling
(Hypothesis~\ref{hypError}) should create large variance among posterior
estimates and cause analyses to be highly sensitive to the number of samples
drawn from the prior.
Furthermore, if insufficient prior sampling is \emph{biasing} estimates toward
models with less parameter space, as suggested by \citet{Hickerson2013}, we
expect to see support for these models decrease as sampling from the prior
increases.
\citet{Oaks2012} did not see such sensitivity when they compared prior sample
sizes of $2\e6$, $5\e6$, and $10^7$.

To explore this prediction further, we repeat the analysis of the Philippines
dataset under the intermediate prior used by \citet{Oaks2012} ($\divt{} \sim
U(0, 10)$, $\meanDescendantTheta{} \sim (0.0005, 0.04)$, $\ancestralTheta{}
\sim (0.0005, 0.02)$), using a very large prior sample size of $10^8$.
When we look at the trace of the estimates of the dispersion index of
divergence times (\vmratio{}) as the prior samples accumulate
(Figure~S\ref{figSamplingError}) we do not see the trend predicted by
Hypothesis~\ref{hypError}.
While approximation error is always present for any numerical analysis, it does
not appear to be playing a large role in the biases revealed by the results of
\citet{Oaks2012} or presented above.

A straightforward prediction if strongly weighted marginal likelihoods are
causing the preference for simple models (Hypothesis~\ref{hypML}) is that the
bias should disappear as the model generating the data converges to the prior.
\citet{Oaks2012} tested this prediction by performing 100,000 simulations to
assess the model-choice behavior of \msb when the prior model is correct.
The results confirm the prediction of Hypothesis~\ref{hypML}:
\msb estimates the probability of the one-divergence model quite well (or even
\emph{under}estimates it) when the prior is correct (see Figure 4 of
\citet{Oaks2012}).
We confirmed this same behavior for the model-averaging approach used by
\citet{Hickerson2013} (see SI text and Figure~S\ref{figValidationMCBehavior}).
These results are not clearly predicted if insufficient computation was causing
numerical error.
Even when the prior is correct, due to the discrete uniform prior on the number
of divergence events (\numt{}) implemented in \msb, models with larger numbers
of divergence-time parameters (and thus greater parameter space) will still be
far less densely sampled than those with fewer divergence events
\citep{Oaks2012}.
Thus, the results of the simulations of \citet{Oaks2012} are more consistent
with the fundamental sensitivity of marginal likelihoods to priors
(Hypothesis~\ref{hypML}).

@@

\section{Improving inference of shared divergences}
In theory, the model-averaging approach of \citet{Hickerson2013} is appealing.
It leverages a great strength of Bayesian statistical procedures, namely the
ability to obtain marginalized estimates that incorporate uncertainty in
nuisance parameters.
However, when sampling over models with narrow-empirical and diffuse uniform
priors for a parameter that is expected to have a very non-uniform likelihood
density, models that exclude the true values of the parameters we aim to
estimate will often have the largest marginal likelihoods.

The recommendations of \citet{Oaks2012} for mitigating the lack of robustness
of \msb are similar to those of \citet{Hickerson2013}, but avoid the need for
imposing an additional dimension of model choice and using priors that often
exclude the truth.
\citet{Oaks2012} suggest that uniform priors may not be ideal for many
parameters of the \msb model, and recommend the use of probability
distributions from the exponential family.
If we look at the prior distribution on divergence times imposed by
the model-averaging approach of \citet{Hickerson2013} we see it is a mixture of
overlapping uniforms with lower limits of zero (Figure~S\ref{figMCTauPrior}).
This looks very much like an exponential distribution, except that in any state
of the model, all the divergence times are restricted to the hard
bounds of one of the uniform distributions.
Thus, it seems more appropriate to simply place a gamma prior (the exponential
being a special case) on divergence times.
This would capture the prior uncertainty that \citet{Hickerson2013} are
suggesting for divergence times (Figure~S\ref{figMCTauPrior}) while avoiding
costly model-averaging and the constraint that all divergence times must fall
within the hard bounds of the current model state.
It also would allow an investigator to place the majority of the prior density
in regions of parameter space they believe, \emph{a priori}, are most
plausible, but still capture uncertainty in the tails of distributions with low
density.
Indeed, \citet{Oaks2014dpp} has shown that the use of gamma distributions in
place of uniform priors improves the power of the method to detect temporal
variation in divergences and reduces erroneous support for clustered
divergences.

\section{Conclusions}
We demonstrate how the approximate Bayesian model-choice method implemented in
\msb can spuriously support models with less parameter space.
This is caused by the use of uniform priors on divergence times.
Uniform distributions necessitate the use of priors that place high density in
unlikely regions of parameter space, less the risk of excluding the true
divergence times \emph{a priori}.
These broad uniform priors reduce the marginal likelihoods of models with more
divergence-time parameters.
We show that the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} does not mitigate this bias, but rather causes it to
manifest by sampling predominantly from models that often exclude the true
values of the divergence times the method is trying to estimate.
Our results show that it is difficult to choose an uniformly distributed prior
on divergence times that is broad enough to confidently contain the true values
of parameters while being narrow enough to avoid strongly weighted and
misleading posterior support for models with less parameter space.
More generally, it is important to carefully choose prior assumptions about
parameters in Bayesian model selection, because they can strongly influence the
\emph{true} posterior probabilities of the models we seek to estimate.

The common inference of temporally clustered historical events
\citep{Barber2010, Bell2012, Carnaval2009, Chan2011, Chan2014, Daza2010,
    Hickerson2006, Huang2011, Lawson2010, Leache2007, Plouviez2009, Stone2012,
    Voje2009},
when not accompanied with the necessary analyses to assess the robustness and
temporal resolution of such results, should be treated with caution, because
\msb has been shown to erroneously infer clustered events over a range of prior
conditions.
Fortunately, \citet{Oaks2014dpp} has shown that alternative probability
distributions allow prior uncertainty to be accommodated while avoiding
excessive prior density in regions of low likelihood, which greatly improves
inference of shared divergence histories.
